---
title: "Explore EMG Normalization"
author: "Lasse Hjorth Madsen"
format: html
editor: visual
---

## Why This?

This note explores match equities in backgammon. Specifically we investigate how EXtremeGammon normalizes match equity calculation to a metric that is (supposedly) comparable to money game equities.

## Introduction

In backgammon money games mistakes are usually evaluated in terms of equities. For example, the non-roller in a pure three-roll position has winning probability, $P(w)$ of:

$$
P(w) = \frac{5}{6}\times\frac{1}{6}\times\frac{5}{6}\ + \ \frac{5}{6}\times\frac{5}{6}\times\frac{5}{6}\times\frac{1}{6} = 0.2122
$$ Since there can't be any propper recubes, the position has an expected value, $E$, of

$$
E= 2P(w) - 1 = -0.5756
$$ for the non-roller. So after accepting an initial double, the non-roller faces an average loss of $2\times-0.6556=-1.1512$ and the position is a money pass, since the loss of one point is less that that.

Suppose this is a match, where the non-roller is trailing 3-away, 2-away sometimes written as (-3,-2). Now, the appropriate unit is *match winning chances*, $mwc$. If doubled, non-roller can take and recube for the match, for a match winning chance of $mwc=0.2122$. Or, 

Link: <https://quarto.org>

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
